---
title: "VO^2^ Max Data"
author: "Andrew Mellor"
date: "06/05/2020"
output:
  html_document:
    toc: true
    toc_depth: 3
---

```{r Setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Overview
This activity will provides the opportunity to practice building multiple linear regression models using R. This activity has been adapted from an excellent resource provided on ^[statistics.laerd.com][statistics.laerd.com](www.statistics.laerd.com). This site provides tutorials that are specific to the statistical software SPSS, however the concepts behind each of the statistical models are still relevant. I recommend checking out this site for more information on a range of different statistical concepts.

## Load required packages

```{r Packages, message=FALSE}
#Include all required packages at the start
library(tidyverse) 
library(ggplot2)
library(broom)
library(readxl)
```

## The Data
The data we will be using for this activity can be found in the vo2-max_data.xlsx file and contains data from 100 participants. It has been sourced from an excellent guide for multiple linear regression on www.statistics.laerd.com and consists of the following variables:

1. subject_id: unique subject identification number  
2. age: subject age in years  
3. weight: subject weight in kilograms  
4. heart_rate: subject average heart rate in the last 5 minutes of a 20 minute submaximal cycling test  
5. gender: 1 = male, 0 = female  
6. vo2_max: VO2max in ml/kg/min recorded from a maximal treadmill test  

Each row is one subjects data.

### Read in the Data

````{R Read in Data}
df <- read_excel("data/vo2-max_data.xlsx")
````
